---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
# Cancer incidences in different NHS health boards for 2019
cancer_data %>% 
  filter(sex == "All" &
           cancer_site == "All cancer types" &
           year == "2019") %>% 
  group_by(hb_name) %>% 
  summarise(all_cancer_incidences = sum(incidences_all_ages)) %>% 
  arrange(desc(all_cancer_incidences))

# Crude rate of cancer in different NHS health boards for 2019
# The crude rate is calculated per 100,000 person-years at risk
cancer_data %>% 
  filter(sex == "All" &
           cancer_site == "All cancer types" &
           year == "2019") %>% 
  group_by(hb_name) %>% 
  summarise(crude_rate_2019 = crude_rate) %>% 
  arrange(desc(crude_rate_2019))

# Incidence rate in NHS Borders by age group, 2015 - 2019
cancer_data_age_groups %>% 
  filter(sex == "All" &
           cancer_site == "All cancer types" &
           hb_name == "NHS Borders") %>% 
  group_by(age_group) %>% 
  summarise(incidence_rate) %>% 
  ggplot(aes(x = reorder(age_group, incidence_rate), y = incidence_rate))+
  geom_col()
```

$$
\text{Crude Rate} = \frac{\text{New Cancer Cases}}{\text{At Risk Population}} * 100,000
$$

```{r}
# Cancer rate over time in Southern Scotland (all cancer types)
cancer_data %>% 
  filter(
    sex == "All" &
    cancer_site == "All cancer types" &
           # & year == "2019"
    hb_name %in% southern_scotland
         ) %>% 
  group_by(hb_name) %>% 
  ggplot(aes(x = year, y = crude_rate, group = hb_name, colour = hb_name))+
  geom_point()+
  geom_line()+
  labs(
    title = "Crude cancer rate over time in Southern Scotland",
    subtitle = "All cancer types",
    x = "\nYear",
    y = "Crude rate\n",
    colour = "Health Board"
  )
```
Analysis

General upward trend over time among all zones on chart.
Possible explanations for this could be:
- increasing size of older population 
- better diagnosis of cancer leading to higher numbers of incidences (useful to
compare to successful treatment rates / life expectancy of cancer patients to see
if these also increase during the time frame)
- as cancer is more prevalent in older age groups, changes in population demographics
(such as younger people moving to more urban areas) could also result in higher rates
but similar incidence numbers

```{r}
# Average rate for top ten cancer types (all Scotland)
cancer_data %>% 
  filter(cancer_site %in% cancer_top_ten) %>% 
  group_by(cancer_site, year) %>% 
  summarise(avg_rate = mean(crude_rate)) %>% 
  arrange(desc(avg_rate)) %>% 
  ggplot(aes(x = year, y = avg_rate, group = cancer_site, colour = cancer_site))+
  geom_point()+
  geom_line()

cancer_top_ten <- cancer_data %>% 
  filter(cancer_site != "All cancer types" 
         & sex == "All"
         & hb_name %in% southern_scotland
         & year == "2019"
         ) %>% 
  group_by(cancer_site) %>% 
  # mutate(total_incidences = sum(incidences_all_ages)) %>% 
  # view()
  summarise(total_incidences = sum(incidences_all_ages), hb_name) %>% 
  arrange(desc(total_incidences)) %>% 
  distinct(cancer_site) %>% 
  head(10) %>% 
  pull()

cancer_top_ten_scotland <- cancer_data %>% 
  filter(cancer_site != "All cancer types" 
         & sex == "All"
         # & hb_name %in% southern_scotland
         & year == "2019"
         ) %>% 
  group_by(cancer_site) %>% 
  # mutate(total_incidences = sum(incidences_all_ages)) %>% 
  # view()
  summarise(total_incidences = sum(incidences_all_ages), hb_name) %>% 
  arrange(desc(total_incidences)) %>% 
  distinct(cancer_site) %>% 
  head(10) %>% 
  pull()

cancer_top_ten_female <- cancer_data %>% 
  filter(cancer_site != "All cancer types" 
         & sex == "Female"
         & hb_name %in% southern_scotland
         & year == "2019"
         ) %>% 
  group_by(cancer_site) %>% 
  # mutate(total_incidences = sum(incidences_all_ages)) %>% 
  # view()
  summarise(total_incidences = sum(incidences_all_ages), hb_name) %>% 
  arrange(desc(total_incidences)) %>% 
  distinct(cancer_site) %>% 
  head(10) %>% 
  pull()

cancer_top_ten_male <- cancer_data %>% 
  filter(cancer_site != "All cancer types" 
         & sex == "Male"
         & hb_name %in% southern_scotland
         & year == "2019"
         ) %>% 
  group_by(cancer_site) %>% 
  # mutate(total_incidences = sum(incidences_all_ages)) %>% 
  # view()
  summarise(total_incidences = sum(incidences_all_ages), hb_name) %>% 
  arrange(desc(total_incidences)) %>% 
  distinct(cancer_site) %>% 
  head(10) %>% 
  pull()

southern_scotland <- c("NHS Borders", "NHS Dumfries and Galloway", "NHS Ayrshire and Arran",
                   "NHS Lanarkshire", "NHS Lothian", "NHS Greater Glasgow and Clyde")

# Average rate for top ten cancer types (Southern Scotland)
cancer_data %>% 
  filter(cancer_site %in% cancer_top_ten &
           hb_name %in% southern_scotland) %>% 
  group_by(cancer_site, year) %>% 
  summarise(avg_rate = mean(crude_rate)) %>% 
  arrange(desc(avg_rate)) %>% 
  ggplot(aes(x = year, y = avg_rate, group = cancer_site, colour = cancer_site))+
  geom_point()+
  geom_line()
```


```{r}
# Changes in top ten cancer types over time in NHS Borders only
cancer_data %>% 
  filter(cancer_site %in% cancer_top_ten &
           hb_name == "NHS Borders" &
           sex == "All") %>% 
  group_by(cancer_site, year) %>% 
  # summarise(avg_rate = mean(crude_rate)) %>% 
  summarise(crude_rate) %>% 
  arrange(desc(crude_rate)) %>%
  ggplot(aes(x = year, y = crude_rate, group = cancer_site, colour = cancer_site))+
  geom_point()+
  geom_line()
```
```{r}
# Changes in top ten cancer types for women over time in NHS Borders only
cancer_data %>% 
  filter(cancer_site %in% cancer_top_ten_female &
           hb_name == "NHS Borders" &
           sex == "Female") %>% 
  group_by(cancer_site, year) %>% 
  # summarise(avg_rate = mean(crude_rate)) %>% 
  summarise(crude_rate) %>% 
  arrange(desc(crude_rate)) %>%
  ggplot(aes(x = year, y = crude_rate, group = cancer_site, colour = cancer_site))+
  geom_point()+
  geom_line()+
    labs(
    title = "Changes in top ten cancer types among women over time",
    subtitle = "NHS Borders",
    x = "\nYear",
    y = "Crude rate\n",
    colour = "Type of cancer\n"
  )
```
```{r}
# Changes in top ten cancer types for men over time in NHS Borders only
cancer_data %>% 
  filter(cancer_site %in% cancer_top_ten_male &
           hb_name == "NHS Borders" &
           sex == "Male") %>% 
  group_by(cancer_site, year) %>% 
  # summarise(avg_rate = mean(crude_rate)) %>% 
  summarise(crude_rate) %>% 
  arrange(desc(crude_rate)) %>%
  ggplot(aes(x = year, y = crude_rate, group = cancer_site, colour = cancer_site))+
  geom_point()+
  geom_line() +
  labs(
    title = "Changes in top ten cancer types among men over time",
    subtitle = "NHS Borders",
    x = "\nYear",
    y = "Crude rate\n",
    colour = "Type of cancer\n"
  )
```


```{r}
cancer_data %>% 
  filter(hb_name == "NHS Borders" &
           cancer_site != "All cancer types") %>% 
  view()
  
```

```{r}
cancer_data_age_groups %>% 
  filter(
    hb_name == "NHS Borders" 
    & cancer_site != "All cancer types"
    & sex == "All"
         ) %>% 
  group_by(age_group) %>% 
  summarise(avg_incidence_rate = mean(incidence_rate)) %>% 
  ggplot(aes(x = age_group, y = avg_incidence_rate))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))+
  labs(
    title = "Average cancer incidence rate by age group",
    subtitle = "2015-19 data for NHS Borders",
    x = "Age group\n",
    y = "\nAverage cancer incidence rate\n",
  )



```

```{r}
# Stacked bar showing prevalence of cancer types in Borders, D&G, Glasgow and Edinburgh (2019)

cancer_data %>% 
  filter(
    hb_name %in% c("NHS Borders", "NHS Dumfries and Galloway", "NHS Lothian", "NHS Greater Glasgow and Clyde")
    & sex == "All"
    & cancer_site %in% cancer_top_ten_scotland
    & year == "2019"
  ) %>% 
  group_by(hb_name, cancer_site) %>% 
  summarise(crude_rate) %>% 
  ggplot(aes(x = reorder(cancer_site, crude_rate), y = crude_rate, group = hb_name, fill = hb_name
             ))+
  geom_bar(position = "dodge", stat = "identity")+
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust = 1))+
  labs(
    title = "Crude rate of top ten cancer types by zone",
    subtitle = "2019 data for Borders, Dumfries & Galloway, Lothians and Greater Glasgow and Clyde",
    x = "\nType of cancer\n",
    y = "Crude rate\n",
    fill = "Health Board"
    )

  

```

