---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
```{r}
library(tidyverse)
library(janitor)
```


Technical Question Example - Cancer Rates in the Borders

Question:

Please prepare a 1 to 2 page analysis summary answering the question below.

In order to help inform the planning for provision of cancer treatment services in NHS Borders, we would like to gain better understanding of the incidence of cancer in NHS Borders.

You will present your findings back to us during the Mock Interviews. Presentations will be 5 minutes long.

```{r}
cancer_data_test <- read_csv("cancer_data_2015_2019.csv") %>% 
  clean_names()

hb_codes <- read_csv("geography_codes_and_labels_hb2014_01042019.csv") %>% 
  clean_names()
```
```{r}
# Join cancer data to health board codes
cancer_data_test_clean <- cancer_data_test %>% 
  left_join(hb_codes, by = "hb") %>% 
  # pivot_longer(
  #   cols = starts_with("incidences"),
  #   names_to = "age_group",
  #   values_to = "incidences"
  # ) %>% 
  # mutate(age_group = str_sub(age_group, start = 12)) %>% 
  
# Pivot longer to create age group and incidence rate variables  
  pivot_longer(
    cols = starts_with("incidence_rate"),
    names_to = "age_group",
    values_to = "incidence_rate"
  ) %>%
# Tidying up new age group column to make more legible  
  mutate(age_group = str_sub(age_group, start = 19),
         age_group = str_replace(age_group, "to", " - "),
         age_group = recode(age_group,
           "_under5" = "0 - 5",
           "85and_over" = "85+"
         )) %>% 
  select(cancer_site, sex, crude_rate, easr, wasr, hb_name, age_group, incidence_rate)
  
```

```{r}
str(cancer_data_clean)
```

Incidence rate incorporates time directly into the denominator and can be calculated as follows:

Number of new cases during a specified time period / person years at risk during the same time period

Person years at risk means the total amount of time (in years) that each person of the study population is at risk of the disease during the period of interest.

```{r}
cancer_data_clean %>% 
  filter(sex == "All" &
           cancer_site == "All cancer types" &
           hb_name == "NHS Borders") %>% 
  group_by(age_group) %>% 
  summarise(incidence_rate) %>% 
  ggplot(aes(x = reorder(age_group, incidence_rate), y = incidence_rate))+
  geom_col()
```

```{r}
cancer_data %>% 
  distinct(year)
```

