---
title: "Cancer rates in the Scottish Borders"
author: "Jonathan Parker"
date: "16/09/2021"
output:
  ioslides_presentation:
        widescreen: true
        css: styles.css
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)

knitr::opts_chunk$set(echo = FALSE,
                          fig.align="center" #align all the figures in the center
                          )

```

```{r, include=FALSE}
library(tidyverse)
library(janitor)

# Read in cancer data and geographic codes and clean names using the Janitor library.
cancer_data <- read_csv("data/cancer_data.csv") %>% 
  clean_names()

cancer_data_age_groups <- read_csv("data/cancer_data_2015_2019.csv") %>% 
  clean_names()

hb_codes <- read_csv("data/geography_codes_and_labels_hb2014_01042019.csv") %>% 
  clean_names()

# Join cancer data to codes and select relevant variables
cancer_data <- cancer_data %>% 
  left_join(hb_codes, by = "hb") %>% 
  select(hb, hb_name, cancer_site, sex, year, incidences_all_ages, crude_rate, easr, wasr)

# Repeat for cancer data with age groups
cancer_data_age_groups <- cancer_data_age_groups %>% 
  left_join(hb_codes, by = "hb") %>% 
# Pivot longer on columns for incidence rates to enable analysis of age groups
  pivot_longer(
    cols = contains("incidence_rate_"),
    names_to = "age_group",
    values_to = "incidence_rate"
  ) %>%
# Tidying up new age group column to make more legible  
  mutate(age_group = str_sub(age_group, start = 19),
         age_group = str_replace(age_group, "to", " - "),
         age_group = recode(age_group,
                            "_under5" = "0 - 4",
                            "5 - 9" = "05 - 09",
                            "85and_over" = "85+"
         )) %>% 
# Selecting relevant variables 
  select(cancer_site, sex, crude_rate, easr, wasr, hb_name, 
         age_group, incidence_rate, incidences_all_ages)

southern_scotland <- c("NHS Borders", "NHS Dumfries and Galloway", "NHS Ayrshire and Arran",
                       "NHS Lanarkshire", "NHS Lothian", "NHS Greater Glasgow and Clyde")

cancer_top_ten <- cancer_data %>% 
  filter(cancer_site != "All cancer types" 
         & sex == "All"
         & hb_name %in% c("NHS Borders", "NHS Dumfries and Galloway", "NHS Ayrshire and Arran",
                   "NHS Lanarkshire", "NHS Lothian", "NHS Greater Glasgow and Clyde")
         & year == "2019"
  ) %>% 
  group_by(cancer_site) %>% 
  # mutate(total_incidences = sum(incidences_all_ages)) %>% 
  # view()
  summarise(total_incidences = sum(incidences_all_ages), hb_name) %>% 
  arrange(desc(total_incidences)) %>% 
  distinct(cancer_site) %>% 
  head(10) %>% 
  pull()

cancer_top_ten_scotland <- cancer_data %>% 
  filter(cancer_site != "All cancer types" 
         & sex == "All"
         # & hb_name %in% southern_scotland
         & year == "2019"
  ) %>% 
  group_by(cancer_site) %>% 
  # mutate(total_incidences = sum(incidences_all_ages)) %>% 
  # view()
  summarise(total_incidences = sum(incidences_all_ages), hb_name) %>% 
  arrange(desc(total_incidences)) %>% 
  distinct(cancer_site) %>% 
  head(10) %>% 
  pull()

cancer_top_ten_female <- cancer_data %>% 
  filter(cancer_site != "All cancer types" 
         & sex == "Female"
         & hb_name %in% c("NHS Borders", "NHS Dumfries and Galloway", "NHS Ayrshire and Arran",
                   "NHS Lanarkshire", "NHS Lothian", "NHS Greater Glasgow and Clyde")
         & year == "2019"
  ) %>% 
  group_by(cancer_site) %>% 
  # mutate(total_incidences = sum(incidences_all_ages)) %>% 
  # view()
  summarise(total_incidences = sum(incidences_all_ages), hb_name) %>% 
  arrange(desc(total_incidences)) %>% 
  distinct(cancer_site) %>% 
  head(10) %>% 
  pull()

cancer_top_ten_male <- cancer_data %>% 
  filter(cancer_site != "All cancer types" 
         & sex == "Male"
         & hb_name %in% c("NHS Borders", "NHS Dumfries and Galloway", "NHS Ayrshire and Arran",
                   "NHS Lanarkshire", "NHS Lothian", "NHS Greater Glasgow and Clyde")
         & year == "2019"
  ) %>% 
  group_by(cancer_site) %>% 
  # mutate(total_incidences = sum(incidences_all_ages)) %>% 
  # view()
  summarise(total_incidences = sum(incidences_all_ages), hb_name) %>% 
  arrange(desc(total_incidences)) %>% 
  distinct(cancer_site) %>% 
  head(10) %>% 
  pull()

```

## Introduction

"In order to help inform the planning for provision of cancer treatment 
services in NHS Borders, we would like to gain better understanding of the 
incidence of cancer in NHS Borders."


- Annual cancer incidence by health board

- Crude rate is calculated as follows:

$$
\text{Crude Rate} = \frac{\text{New Cancer Cases}}{\text{At Risk Population}} * 100,000
$$

## Cancer rate over time in Southern Scotland
```{r, fig.width=9,fig.height=5, warning=FALSE}
# Cancer rate over time in Southern Scotland (all cancer types)
cancer_data %>% 
  filter(
    sex == "All" &
    cancer_site == "All cancer types" &
    hb_name %in% southern_scotland
         ) %>% 
  group_by(hb_name) %>% 
  ggplot(aes(x = year, y = crude_rate, group = hb_name, colour = hb_name))+
  geom_point()+
  geom_line()+
  labs(
    title = "Crude cancer rate over time in Southern Scotland",
    subtitle = "All cancer types",
    x = "\nYear",
    y = "Crude rate\n",
    colour = "Health Board"
  )
```


## Top ten cancer types for women, NHS Borders
```{r, fig.width=9,fig.height=5, warning=FALSE}
# Changes in top ten cancer types for women over time in NHS Borders only
cancer_data %>% 
  filter(cancer_site %in% cancer_top_ten_female &
           hb_name == "NHS Borders" &
           sex == "Female") %>% 
  group_by(cancer_site, year) %>% 
  summarise(crude_rate) %>% 
  arrange(desc(crude_rate)) %>%
  ggplot(aes(x = year, y = crude_rate, group = cancer_site, colour = cancer_site))+
  geom_point()+
  geom_line()+
    labs(
    title = "Changes in top ten cancer types among women over time",
    subtitle = "NHS Borders",
    x = "\nYear",
    y = "Crude rate\n",
    colour = "Type of cancer\n"
  )
```

## Top ten cancer types for men, NHS Borders
```{r, fig.width=9,fig.height=5, warning=FALSE}
cancer_data %>% 
  filter(cancer_site %in% cancer_top_ten_male &
           hb_name == "NHS Borders" &
           sex == "Male") %>% 
  group_by(cancer_site, year) %>% 
  # summarise(avg_rate = mean(crude_rate)) %>% 
  summarise(crude_rate) %>% 
  arrange(desc(crude_rate)) %>%
  ggplot(aes(x = year, y = crude_rate, group = cancer_site, colour = cancer_site))+
  geom_point()+
  geom_line() +
  labs(
    title = "Changes in top ten cancer types among men over time",
    subtitle = "NHS Borders",
    x = "\nYear",
    y = "Crude rate\n",
    colour = "Type of cancer\n"
  )
```

## Rural vs Urban
```{r, fig.width=9,fig.height=5, warning=FALSE}
cancer_data %>% 
  filter(
    hb_name %in% c("NHS Borders", "NHS Dumfries and Galloway", "NHS Lothian", "NHS Greater Glasgow and Clyde")
    & sex == "All"
    & cancer_site %in% cancer_top_ten_scotland
    & year == "2019"
  ) %>% 
  group_by(hb_name, cancer_site) %>% 
  summarise(crude_rate) %>% 
  ggplot(aes(x = reorder(cancer_site, crude_rate), y = crude_rate, group = hb_name, fill = hb_name
             ))+
  geom_bar(position = "dodge", stat = "identity")+
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust = 1))+
  labs(
    title = "Crude rate of top ten cancer types by zone",
    subtitle = "2019 data for Borders, Dumfries & Galloway, Lothians and Greater Glasgow and Clyde",
    x = "\nType of cancer\n",
    y = "Crude rate\n",
    fill = "Health Board"
    )
```

## Key findings

1. Increase in cancer rate is likely due to increasing size of older population

2. Higher rate in rural areas can also be explained differing population demographics

3. Skin cancers seem to have the highest rates among both men and women in NHS Borders

### Recommendation

- Focus more efforts on identification and treatment of skin cancers



