---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
library(tidyverse)
library(GGally)
library(fastDummies)
library(janitor)
library(ggfortify)

diamonds <- read_csv("data/diamonds.csv")
```

```{r}
summary(diamonds)
```
We expect the carat of the diamonds to be strong correlated with the physical dimensions x, y and z. Use ggpairs() to investigate correlations between these four variables.

```{r}
diamonds %>% 
  select(carat, x, y, z) %>% 
  ggpairs()
```

So, we do find significant correlations. Let’s drop columns x, y and z from the dataset, in preparation to use only carat going forward.

```{r}
diamonds_trim <- diamonds %>% 
  select(-c("x", "y", "z"))
```


We are interested in developing a regression model for the price of a diamond in terms of the possible predictor variables in the dataset.

Use ggpairs() to investigate correlations between price and the predictors (this may take a while to run, don’t worry, make coffee or something).

Perform further ggplot visualisations of any significant correlations you find.

```{r}
ggpairs(diamonds_trim)
```

```{r}
diamonds_trim %>% 
  ggplot(aes(x = carat, y = price))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
diamonds_trim %>% 
  ggplot(aes(x = cut, y = price))+
  geom_boxplot()
```

```{r}
diamonds_trim %>% 
  ggplot(aes(x = clarity, y = price))+
  geom_boxplot()
```

```{r}
diamonds_trim %>% 
  dummy_cols(select_columns = "cut", remove_selected_columns = TRUE,
                          remove_first_dummy = TRUE) %>% 
  clean_names()
```

```{r}
diamonds_trim %>% 
  dummy_cols(select_columns = "color", remove_selected_columns = TRUE,
                          remove_first_dummy = TRUE) %>% 
  clean_names()
```

```{r}
diamonds_trim %>% 
  dummy_cols(select_columns = "clarity", remove_selected_columns = TRUE,
                          remove_first_dummy = TRUE) %>% 
  clean_names()
```

```{r}
model <- lm(price ~ carat, data = diamonds_trim)
```
```{r}
autoplot(model)
```
```{r}
summary(model)
```

```{r}
model2 <- lm(price ~ carat + clarity, data = diamonds_trim)
```
```{r}
autoplot(model2)
```
```{r}
summary(model2)
```
