---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
library(gtools)
library(prob)
library(tidyverse)
```
```{r}
titanic <- read_csv("data/titanic.csv")
```
MVP
1.1

A data set detailing the survival or death of passengers and crew members on the RMS Titanic is given in the data folder for this homework - use this data and not the built-in Titanic dataset in R.

Some time after the disaster a journalist wrote that crew members were more likely to have survived the sinking than passengers. This was disputed by a descendant of one of the surviving crew members, who suggested that the opposite was true, i.e. passengers were more likely to have survived than crew.

Question
Calculate the conditional probabilities prob(survived∣crew) and prob(survived∣passenger)
to settle this dispute.

Have a look at the distinct values of the class variable, this will tell you whether each individual was a passenger (1st, 2nd or 3rd class) or a member of the crew (treat restaurant staff as crew)

```{r}
titanic %>% 
  distinct(class)

passenger <- c("1st", "2nd", "3rd")

passengers_survived <- titanic %>% 
  mutate(passenger = if_else(class %in% passenger, TRUE, FALSE)) %>% 
  filter(survived == "yes",
         passenger == TRUE) %>% 
  group_by(passenger) %>% 
  summarise(count = n()) %>% 
  pull()

passengers_died <- titanic %>% 
  mutate(passenger = if_else(class %in% passenger, TRUE, FALSE)) %>% 
  filter(survived == "no",
         passenger == TRUE) %>% 
  group_by(passenger) %>% 
  summarise(count = n()) %>% 
  pull()

crew_survived <- titanic %>% 
  mutate(passenger = if_else(class %in% passenger, TRUE, FALSE)) %>% 
  filter(survived == "yes",
         passenger == FALSE) %>% 
  group_by(passenger) %>% 
  summarise(count = n()) %>% 
  pull()

crew_died <- titanic %>% 
  mutate(passenger = if_else(class %in% passenger, TRUE, FALSE)) %>% 
  filter(survived == "no",
         passenger == FALSE) %>% 
  group_by(passenger) %>% 
  summarise(count = n()) %>% 
  pull()

total_passengers <- passengers_survived + passengers_died
total_crew <- crew_survived + crew_died

prob_survived_passenger <- passengers_survived / total_passengers
prob_survived_crew <- crew_survived / total_crew

# Passengers' probability of surviving: 38%
# Crew probability of surviving: 24%

prob_survived_passenger
prob_survived_crew

# ANSWER
titanic_probs <- titanic %>%
  group_by(crew) %>%
  summarise(prob = sum(survived == "yes") / n()) 

titanic_probs
```





1.2 Probability Consolidation Questions

Today was pretty full-on! Here are some problems to help you consolidate what you learned. Feel free to work on paper or in R, whichever you are most comfortable with:

MVP - try four of the problems
Extension - try more than four!


Question 1.
If I roll a six-sided dice once, what is the probability of getting a number less than 5?
```{r}
s <- rolldie(1) %>%
  nrow()

s_one_dice <- rolldie(1) %>% 
  filter(X1 < 5) %>% 
  summarise(count = n()) %>% 
  pull()

prob_less_than_5 <- s_one_dice / s

prob_less_than_5
```


Question 2.
How many outcomes are there in the sample space of an experiment in which we toss three dice?
```{r}
three_dice <- rolldie(3) %>% 
  nrow()

three_dice
```


Question 3.
In an experiment where we roll four dice:

    how many outcomes are in the event where the total of the four dice equals 5, 6 or 7?
    what is the probability of this event?
    what assumptions do we apply in answering part ii?

Hint
For part i, rowSums(.) may help you setup up the necessary random variable total (or you can use pivot_longer()) 

```{r}
#     how many outcomes are in the event where the total of the four dice equals 5, 6 or 7?
four_dice_567 <- rolldie(4) %>% 
  mutate(sum = rowSums(.)) %>% 
  filter(sum == 5 | sum == 6 | sum == 7) %>% 
  nrow()

four_dice <- rolldie(4) %>% 
  nrow()

prob_four_dice_567 <- four_dice_567 / four_dice

#     what is the probability of this event?
prob_four_dice_567

#    what assumptions do we apply in answering part ii?
#    ???

```



Question 4.
Five chairs are placed in a row in a room, along with five people of differing height. The people sit down randomly on the chairs. What is the probability that they sit in order of decreasing height, left to right?
Hint
This is an arrangement where order matters. Of all the possible arrangements, how many are correct?

```{r}

heights <- c(1.6, 1.7, 1.8, 1.9, 2)

chairs <- as_tibble(
  permutations(
    n = 5,
    r = 5,
    v = heights
  )
)

decr_height <- chairs %>% 
  filter(V1 == 2 & V2 == 1.9 & V3 == 1.8 & V4 == 1.7) %>% 
  nrow()

prob_order_decr_height <- decr_height / nrow(chairs)
prob_order_decr_height


```


Question 5.
I roll six dice:

    what is the probability I see the numbers 6, 5, 4, 3, 2 and 1 in order on the dice, left to right?
    what is the probability I see one each of the numbers 6, 5, 4, 3, 2 and 1 on the dice regardless of order?

Hint
Have a look at the function in the package. For example, to all rows containing , , in that order, you could use

```{r}
options(scipen=999)
roll_order <- rolldie(6) %>% 
  filter(X1 == 6 & X2 == 5 & X3 == 4 & X4 == 3 & X5 == 2 & X6 == 1) %>% 
  nrow()

#     what is the probability I see the numbers 6, 5, 4, 3, 2 and 1 in order on the dice, left to right?
prob_roll_order_six_one <- roll_order / nrow(rolldie(6))
prob_roll_order_six_one

#    what is the probability I see one each of the numbers 6, 5, 4, 3, 2 and 1 on the dice regardless of order?
roll_no_order <- rolldie(6) %>% 
  mutate(sum = rowSums(.)) %>% 
  filter(sum == sum(1:6)) %>% 
  nrow()

prob_roll_no_order_six_one <- roll_no_order / nrow(rolldie(6))
prob_roll_no_order_six_one

# ANSWER

contains_654321 <- six_rolls %>%
  filter(isin(., c(6, 5, 4, 3, 2, 1), ordered = FALSE))

prob <- nrow(contains_654321) / nrow(six_rolls)
prob


```


Question 6.
If I draw a single card from a 52-card deck, and it is black, what is the probability that its suit is spades?
Hints

```{r}
black <- cards(jokers = FALSE) %>% 
  filter(suit == "Club" | suit == "Spade") %>%
  nrow()

spades <- cards(jokers = FALSE) %>% 
  filter(suit == "Spade") %>% 
  nrow()

prob_spade <- spades / black
prob_spade
  
```


2 Extension

Answer more than 4 consolidation questions
