---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
1 MVP

Now we’ll go back to CI creation in the normal fashion. We’ll take the ames data from the CIs lab earlier today and regard it now as a sample, we won’t be drawing any smaller samples from within it. This is the usual situation in an analysis: you use all the data available to you!

Task 1.
Load the data again, clean_names(), and re-familiarise yourself with it

```{r}
library(janitor)
library(tidyverse)
library(infer)

ames <- read_csv("data/ames.csv") %>% 
  clean_names()
```

Task 2.
Investigate the distribution of lot_area. Is the distribution roughly normal? If not, what problems do you find?
```{r}

lot_area_histo <- ames %>% 
  ggplot(aes(x = lot_area)) +
  geom_histogram(col = "white") +
  xlim(0, 50000)

lot_area_box <- ames %>% 
  ggplot(aes(x = lot_area)) +
  geom_boxplot()

library(e1071)

ames %>%
  summarise(skewness = skewness(lot_area, type = 1))

# Distribution is quite normal below 30,000 but highly right-skewed due to
# outliers.
```

Task 3.
Compute and visualise a bootstrap sampling distribution for the mean(lot_area) of the sold houses.
```{r}

bootstrap_sample <- ames %>% 
  rep_sample_n(size = 200, replace = TRUE, reps = 5000) %>% 
  summarise(mean_lot_area = mean(lot_area))

bootstrap_sample_plot <- bootstrap_sample %>% 
  ggplot(aes(x = mean_lot_area)) +
  geom_histogram(col = "white")

infer_bootstrap_sample <- ames %>% 
  specify(response = lot_area) %>% 
  generate(reps = 5000, type = "bootstrap") %>% 
  calculate(stat = "mean")

```

Task 4.
Use your bootstrap distribution to calculate a 95% CI for mean(lot_area), and visualise it on the distribution

```{r}
ci_95 <- infer_bootstrap_sample %>% 
    get_ci(level = 0.95, type = "percentile")

ci_95

#     mean lower_bound upper_bound
# 1 10154.       9280.      11454.

infer_bootstrap_sample %>% 
  visualise(method = "simulation") +
  shade_ci(endpoints = ci_95)
```

Task 5.
You would like to know the mean(lot_area) of the sold houses with higher confidence. Calculate the 99% CI for this variable (you can re-use your bootstrap distribution from above). Is it narrower or broader than the 95% CI? Does that make sense?

```{r}
ci_99 <- infer_bootstrap_sample %>% 
  get_ci(level = 0.99, type = "percentile")

infer_bootstrap_sample %>% 
  visualise(method = "simulation") +
  shade_ci(endpoints = ci_99)

# Broader distribution than the 95% CI, which is expected.
```

Task 6.
Calculate the point estimate of the mean(lot_area)
```{r}
summary_infer_bootstrap_sample <- infer_bootstrap_sample %>% 
  ungroup() %>% 
  summarise(
    mean_lot_area = mean(stat)
  )

summary_infer_bootstrap_sample
```

